paper2_resistance_mechanisms_mixtures

### 2017-03-24 inst/documents/paper2/paper2_resistance_mechanisms_mixtures.Rmd
paper2 writing

The model outputs the change in resistance allele frequencies over time (generations). We use this to calculate a time-to-resistance using a resistance threshold of 50%. This is consistent with thersholds used in previous modelling analyses [todo refs] and represents a level where the resistance alleles are close to fixation.

@@

1. single insecticide
1. two insecticides used in sequence, replacing the first with a second once the proportion of resistance alleles reaches a threshold of 50%
1. two insecticides used in a mixture (concurrently)

@@

When single insecticide use was investigated (Figs 3-4) higher values of inputs that increased the selective advantage of resistance all led to more rapid evolution resistance, and higher values for cost led to slower evolution of (Fig 4B). The latter is as expected because it reduces the selective advantage of resistance.

For single insecticide use resistance responded identically to changing exposure and effectiveness (compare Fig 3A to 3B). This makes sense as vector kill is effectively a product of exposure times effectiveness. For example exposing 50% of a  population to an insecticide which is 75% effective would be expected to have the same result as exposing 75% of the population to one which is 50% effective. This observation points to the mechanism by which increasing both exposure and effectiveness lead to a faster increase in resistance. In both cases the increased deaths of susceptible vectors cause a higher selection pressure that can explain the faster increase.

The pattern of more rapid increase in resistance at higher dominance of restoration levels (Fig 3C) can be explained by increased survival of heterozygotes in the presence of the insecticide. Higher dominance of the resistant allele causes it to contribute more to the phenotype of the heterozygotes leading to higher survival. Thus selection pressure for the resistance allele will be increased because it confers more of an advantage when only present on one chromosome. The faster development of resistance under higher resistance-restoration (Fig 3D) can be explained by it's effect on the survival of the resistant genotypes. Resistance-restoration restores the survival of resistant genotypes in the presence of the insecticide back towards what it would be in the absence of the insecticide, thus increasing the selective advantage of the resistance allele. The effect of the starting frequency of resistance (Fig 4) is the most different from the other inputs. Simply, when starting from a higher frequency of resistance there is a smaller change to make to reach the higher resistance thresholds.

These mechanisms, explaining the response of resistance to single insecticide use under different input values, are summarised in Table 2. Exposure and effectiveness act by decreasing the survival of all genotypes, dominance of restoration acts by increasing survival of RS and resistance-restoration acts by increasing survival of RR & RS.

#### Two insecticides
Using two insecticides either in sequence or in mixtures differs from the use of a single insecticide and, where it differs can again help us to understand the dynamics of resistance evolution. Our base scenario (Fig 5A) shows curves for resistance to both insecticides in the mixture that are shallower than those for sequential use. The shallower curves for the mixture can be explained by each insecticide killing individuals that are resistant to the other insecticide. Thus each insecticide reduces the selection pressure for the increase in resistance to the other and could be said to 'protect' the other. However this protection is not necessarily enough to ensure that resistance arises more slowly in a mixture relative to a sequential strategy. Resistance can evolve more slowly for both insecticides in sequential use, despite being faster for each, because they occur one after the other. In Figure 5A it takes 60 generations for resistance to reach the threshold for each insecticide in the sequence, and 90 when they are applied together in the mixture.

#### Insecticide effectiveness

The main difference between mixtures and the other strategies is that in mixtures increasing the effectiveness of either insecticide slows the development of resistance to both (compare Fig 5B to 5A). In contrast in both the sequential (Fig 5B) and single insecticide (Fig 2A) scenarios the development of resistance was faster when effectiveness was higher. This can be explained by the increased effectiveness of one insecticide in a mixture increasing killing of vectors that are resistant to the other and thus decreasing selection pressure in a protective effect.

This effect can be seen by comparing Fig 5B, where the effectiveness of insecticide 1 has been increased, to Fig 5A. In the mixture, resistance to insecticide 1 (with the increased effectiveness) rises at a similar speed in both figures. However, the increased effectiveness of insecticide 1 causes resistance to insecticide 2 to increase more slowly initially in the mixture. Once insecticide 1 reaches resistance of around 50% at around 50 generations, the curve for insecticide 2 becomes steeper. Once the first insecticide has become ineffective due to resistance it stops killing individuals that are resistant to the second insecticide, loses its protective effect and thus stops slowing the rise in resistance to the second insecticide. A similar effect is visible when the effectiveness of both insecticides are increased (Fig 6C). The identical resistance curves for each insecticide in the mixture are shallower than at the lower effectiveness (Fig 5A) because more individuals resistant to each insecticide are being killed by the other insecticide. In this case the 'protection' given to both insecticides by the other declines at the same rate so we don't see the change in slope seen in Fig 5B.

@@

#### Difference between exposure and effectiveness
It seems initially counter-intuitive that whereas increasing effectiveness slows the development of resistance for mixtures, increasing exposure speeds it up. Increasing the exposure to both insecticides results in a decrease in time-to-resistance (from 90 generations in Fig 5A to 40 generations in Fig 5C) where increasing the effectiveness of both insecticides by the same amount results in a slight increase (from 90 in Fig 5A to 110 in Fig 5B). This contrasts with the identical effects that exposure and effectiveness have on a single insecticide (Figs 3 A,B) and on insecticides in sequence (the red dashed lines are the same in Figs 5B & 5C). This difference between effectiveness and exposure, that effectiveness increases times-to-resistance for a mixture and exposure reduces it, was unexpected for us.

Effectiveness and exposure were also found to be the most important inputs in our earlier analysis where we conducted a more comprehensive sensiitivty analysis [@Levick2017]. If these predicted effects are confirmed then in designs of insecticide mixtures it would be best in terms of evolution to have high effectiveness and low exposure. This low exposure could be achieved by having active ingredients with some repellency.

#### Costs of resistance
Costs of resistance favour mixtures in these analyses. This is because the evolution of a resistance with an associated cost is slowed more in a mixture than in a sequence (compare the red solid lines for a mixture to the red dotted line for a sequence in Fig 9C and D). In a mixture the protective effect of the other insecticide identified above interacts with the cost of resistance to reduce selection pressure further.

It can also be seen that the effects of cost of resistance become more as the frequency of resistance increases. The resistance curves for two insecticides in a mixture that differ only in their cost of resistance, diverge more as the frequency of resistance increases (Fig 9C). This makes intuitive sense, as the frequency of resistance genes increases the impact of fitness costs associated with those genes becomes greater. Thus differences due to cost are less noticeable at low resistance frequencies and become more noticeable at higher resistance frequencies. Thus if a lower resistance threshold was used cost would have less of an effect and if a higher resistance threshold was used cost would have more of an effect. Also note that if the costs are set even higher than we set them here (e.g. above 0.2) they can maintain resistance frequencies below 1 or lead to their decline. Unfortunately, we do not expect such high costs to occur in operational settings.

However, also be aware that in our analyses we have used a single resistance threshold. When there are costs of resistance, the frequency of resistance for insecticides that are not being used is likely to decline. Thus the frequency of resistance for the first insecticide in a sequence may decline to a level where it becomes operationally useful again. As such a repeated sequential strategy could offer benefits over mixtures.

#### Caveats
The analysis presented here examines solely the evolution of resistance and not the control of mosquitoes or malaria. Different insecticide use strategies, as well as effecting the evolution of resistance, will also influence both the numbers of vectors and their propensity to transmit the disease [@Rivero2010]. Insecticide resistance could reduce vector transmission of disease (e.g. by reduced lifespan of the vector) or increase transmission (e.g. by reduced immunity of the vector to the disease) [@Rivero2010]. Our model would predict that the best strategy for limiting the development of resistance in the short term would be to use no insecticide. Of course, using no insecticide could have serious consequences for mosquito and malaria abundance. The challenge is to develop good strategies for delivering mosquito and malaria control in the short term while sustaining their operational capacity in the longer term by restricting the development of resistance. Others are combining models of population genetics and population dynamics to investigate these trade-offs [@Barbosa2017].

@@

The development of insecticide resistance in operational field settings is subject to considerable uncertainty [@Ranson2016]. The model we present here is able to cover the range of that uncertainty, yet in this paper, to develop an understanding of the key mechanisms we restrict our explorations to a limited region of parameter space (e.g. holding most inputs constant at intermediate values while modifying single inputs in isolation). Thus there is the potential that the system may behave differently in the field than we describe here. However, the results presented here are supported by an earlier analysis where we ran 10,000 unique scenarios varying more inputs [@Levick2017]. This previous analysis also highlighted the relative importance of insecticide effectiveness and exposure in determining the relative performance of mixtures and sequences. In addition we provide an on-line user interface to the model [@South2017a] and enabling readers to investigate model behaviour in areas of parameter space we have not shown here.

#### Putting results into context

The predicted responses of resistance presented here can be put into the context of previous work and recommendations. Earlier modelling work has made sometimes contrasting predictions and these can be explained by the details of their structure and input choices. One model predicted that "the use of mixtures is always more effective in delaying the onset of resistance often by many orders of magnitude"[@Mani1985]. That model assumed that SS genotypes were always killed, RR always survived, and that a constant 10% of mosquitoes escaped the insecticide. This would be represented in our model by a very high 'insecticide effectiveness' of 1, a 'resistance restoration' of 1 and an 'exposure' of 0.9. This result is consistent with our prediction that the very high effectiveness is likely to favour mixtures over sequences (e.g. see Fig 6D and Table 4).

A subsequent model [@Roush1998] lead to the following more pessimistic conclusion about the potential of mixtures : “As a result of incomplete coverage and residue decay, the mortality of susceptible homozygotes is rarely consistently high enough for pesticide mixtures to be effective” [@Roush1998]. This "mortality of susceptible homozygotes" is equivalent to the parameter we have termed insecticide effectiveness, so lends further support to our main conclusion of the importance of effectiveness. It's also shown [@Roush1998 Fig 3] that decreasing exposure from 90% to 80% also increases the benefit of mixtures over sequences in the same way that we have shown.

One consistent aspect of the results presented here can at first seem to contradict published advice. We show that lower insecticide effectiveness or exposure are expected to lead to slower spread of resistance in sole use and sequences (Figs 3 & 5, Table 2). This is consistent with expectations that lower effectiveness and exposure would lead to lower mortality of susceptibles and thus lower selection pressure for resistance [e.g. @Roush1998]. In contrast the widely held view seems to be that poor implementation can promote the development of resistance. e.g. from [@FAO2012] "Certain pest control practices have consistently been shown to exacerbate the loss of susceptible pest populations and the development of resistance. These include: ... - the use of application rates that are below or above those recommended on the label, - poor coverage of the area being treated ...". The explanation for this apparent difference is in the effect of dominance. Poor application can reduce the mortality of heterozygotes (RS) thus effectively increasing the dominance of the resistant allele and increasing its rate of evolution [@Barbosa2011][@FAO2012]. Our model results are consistent with this, showing that increasing dominance of restoration leads to faster spread of resistance (Fig 3C, 7). So, the effect of poor implementation of an insecticide intervention on resistance could be either positive or negative depending on the relative effects on dominance of the resistance gene, insecticide effectiveness and exposure. Similarly, the effects of poor application on the relative benefits of mixtures and sequences will depend on these trade-offs because of our result that effectivenes favours mixtures, exposure favours sequences and higher dominance favours neither (Table 4).

@@

#### Future work

This model has been designed to be flexible to address further issues in Insecticide Resistance Management. With a little restructuring our model can represent other insecticide use strategies such as rotations and mosaics.

#### FINAL CONCLUDING PARAGRAPH

(@) Effectiveness and exposure are key and can be modified. Cost also has an effect and future work may want to look more at it's interaction with these other important inputs.

(@) In a mixture each insecticide reduces the rate of increase in resistance to the other by killing individuals that are resistant to the other.

(@) In a mixture higher effectiveness of either insecticide kills more individuals resistant to the other and thus increases time-to-resistance for the other. Thus although increasing effectiveness of an insecticide decreases time-to-resistance when used alone, when used in a mixture time-to-resistance for both insecticides is increased.

(@) For a mixture of 2 insecticides with differing effectiveness. Resistance to the insecticide with a greater effectiveness increases faster. The less effective insecticide is 'protected' by the more effective. Resistance to the less effective insecticide increases slowly until resistance to the more effective insecticide reaches a high level.

(@) The model we describe is flexible and as such can reproduce similar results to earlier models and be used to represent more detailed, operationally relevant strategies.

(@) This model can explore when a mixture strategy may be favoured over a sequence purely for the evolution of resistance. Other factors also need to be considered. On the positive side a mixture is likely to kill more vectors and may also limit their ability to transmit the disease. On the negative side a mixture is more expensive.

>>>>

The model outputs the change in resistance allele frequencies over time measured in generations. We extract the number of generations at which a resistance allele frequency of 50% is reached and term this 'time-to-resistance'. Using a 50% threshold is consistent with previous modelling analyses e.g. [@Curtis1985][@Birget2015]. Preliminary investigations showed that using other potential thresholds of 25 or 75% gave very similar results.

@@

1. single insecticide
1. two insecticides used in sequence, replacing the first with a second once the frequency of resistance alleles reaches a threshold of 50%
1. two insecticides used in a mixture (concurrently)

@@

When single insecticide use was investigated (Figs 3-4) higher values of inputs that increased the selective advantage of resistance all led to more rapid evolution resistance, and higher values for cost led to slower evolution of (Fig 4B).

These mechanisms, explaining the response of resistance to single insecticide use under different input values, are summarised in Table 2. Referring to figure 1 can help explain these. Exposure and effectiveness act by decreasing the fitness of all genotypes, dominance of restoration acts by increasing fitness of SR and resistance-restoration acts by increasing fitness of RR & SR. Cost acts by decreasing the fitness of RR and SR for those not exposed.

#### Two insecticides
Using two insecticides either in sequence or in mixtures differs from the use of a single insecticide and, where it differs can again help us to understand the dynamics of resistance evolution. Our base scenario (Fig 5A) shows curves for resistance to both insecticides in the mixture that are shallower than those for sequential use. The shallower curves for the mixture can be explained by each insecticide killing individuals that are resistant to the other insecticide. Thus each insecticide reduces the selection pressure for the increase in resistance to the other and could be said to 'protect' the other. However this protection is not necessarily enough to ensure that resistance arises more slowly in a mixture relative to a sequential strategy. Resistance can evolve more slowly for both insecticides in sequential use, despite being faster for each, because they occur one after the other. In Figure 5A it takes 60 generations for resistance to reach the threshold for each insecticide in the sequence, and 90 when they are applied together in the mixture, so the sequence is still favoured because 2x60 is greater than 90. Incraesing the effectiveness of on insecticide to Fig 5B decreases the added times for the sequence to 90 and increases those for the mixture to 115 thus changing the order and favouring the mixture.

The mechanism for this change can be sen by comparing the shape of the curves in Figs 5A & B. In the mixture, resistance to insecticide 1 (with the increased effectiveness) rises at a similar speed in both figures. However, the increased effectiveness of insecticide 1 causes resistance to insecticide 2 to increase more slowly initially in the mixture. Once insecticide 1 reaches resistance of around 50% at around 50 generations, the curve for insecticide 2 becomes steeper. Once the first insecticide has become ineffective due to resistance it stops killing individuals that are resistant to the second insecticide, loses its protective effect and thus stops slowing the rise in resistance to the second insecticide. A similar effect is visible when the effectiveness of both insecticides are increased (Fig 6C). The identical resistance curves for each insecticide in the mixture are shallower than at the lower effectiveness (Fig 5A) because more individuals resistant to each insecticide are being killed by the other insecticide. In this case the 'protection' given to both insecticides by the other declines at the same rate so we don't see the change in slope seen in Fig 5B.

@@

#### Difference between exposure and effectiveness
It seems initially counter-intuitive that whereas increasing effectiveness slows the development of resistance for mixtures, increasing exposure speeds it up. Increasing the exposure to both insecticides results in a decrease in time-to-resistance (from 90 generations in Fig 5A to 40 generations in Fig 5C) where increasing the effectiveness of both insecticides by the same amount results in a slight increase (from 90 in Fig 5A to 115 in Fig 5B). This contrasts with the identical effects that exposure and effectiveness have on a single insecticide (Figs 3 A,B) and on insecticides in sequence (the red dashed lines are the same in Figs 5B & 5C). This difference, that exposure reduces times-to-resistance for a mixture and effectiveness increases it, was unexpected for us.

#### Costs of resistance
Costs of resistance favour mixtures in these analyses because the evolution of resistance is slowed more by cost in a mixture than in a sequence (compare the red solid lines for a mixture to the red dotted line for a sequence in Fig 9C and D). It seems that the protective effect of the other insecticide in a mixture interacts with the cost of resistance to reduce selection pressure more than in a sequence.

The effects of cost of resistance become more as the frequency of resistance increases. The resistance curves for two insecticides in a mixture that differ only in their cost of resistance, diverge more as the frequency of resistance increases (Fig 9C). This makes intuitive sense, as the frequency of resistance genes increases the impact of fitness costs associated with those genes becomes greater. Thus differences due to cost are less noticeable at low resistance frequencies and become more noticeable at higher resistance frequencies. Thus if a lower resistance threshold was used cost would have less of an effect and if a higher resistance threshold was used cost would have more of an effect. Also note that if the costs are set even higher than we set them here (e.g. above 0.2) they can maintain resistance frequencies below 1 or lead to their decline. Unfortunately, we do not expect such high costs to occur in operational settings.

Relatedly, when there are costs of resistance, the frequency of resistance for insecticides that are not being used is likely to decline. Thus the frequency of resistance for the first insecticide in a sequence, after it has been replaced, may decline to a level where it becomes operationally useful again. As such a repeated sequential strategy could offer benefits over mixtures not considered here. To incorporat this one would need to use a more detailed metric than time-to-resistance.

#### Caveats
Our model examines solely the evolution of resistance and not the control of mosquitoes or malaria. Different insecticide use strategies, as well as effecting the evolution of resistance, will also influence both the numbers of vectors and their propensity to transmit the disease [@Rivero2010]. Insecticide resistance could reduce vector transmission of disease (e.g. by reduced lifespan of the vector) or increase transmission (e.g. by reduced immunity of the vector to the disease) [@Rivero2010]. Our model would predict that the best strategy for limiting the development of resistance in the short term would be to use no insecticide. Of course, using no insecticide could have serious consequences for mosquito and malaria abundance. The challenge is to develop good strategies for delivering mosquito and malaria control in the short term while sustaining their operational capacity in the longer term by restricting the development of resistance. Others are combining models of population genetics and population dynamics to investigate these trade-offs [@Barbosa2017].

@@

The development of insecticide resistance in operational field settings is subject to considerable uncertainty [@Ranson2016]. The model we present here is able to cover the range of that uncertainty, yet in this paper, to develop an understanding of the key mechanisms we restrict our explorations to a limited region of parameter space (e.g. holding most inputs constant at intermediate values while modifying single inputs in isolation). Thus there is the potential that the system may behave differently in the field than we describe here. However, the results presented here are supported by an earlier analysis where we ran 10,000 unique scenarios varying more inputs [@Levick2017]. This previous analysis also highlighted the relative importance of insecticide effectiveness and exposure in determining the relative performance of mixtures and sequences. In addition we provide an on-line user interface to the model [@South2017a] and enabling readers to investigate model behaviour in areas of parameter space we have not shown.

#### Consistency with previous work 
The predicted responses of resistance presented here can be put into the context of previous work and recommendations. Earlier modelling work has made sometimes contrasting predictions and these can be explained by the details of their structure and input choices. One model predicted that "the use of mixtures is always more effective in delaying the onset of resistance often by many orders of magnitude"[@Mani1985]. That model assumed that SS genotypes were always killed, RR always survived, and that a constant 10% of mosquitoes escaped the insecticide. This would be represented in our model by a very high 'insecticide effectiveness' of 1, a 'resistance restoration' of 1 and an 'exposure' of 0.9. This result is consistent with our prediction that the very high effectiveness is likely to favour mixtures over sequences (e.g. see Fig 6D and Table 4).

A subsequent model [@Roush1998] lead to the following more pessimistic conclusion about the potential of mixtures : “As a result of incomplete coverage and residue decay, the mortality of susceptible homozygotes is rarely consistently high enough for pesticide mixtures to be effective” [@Roush1998]. This "mortality of susceptible homozygotes" is equivalent to our insecticide effectiveness. Again this points to the importance of effectiveness and the question of whether the effectiveness levels needed to make mixtures better are attainable operationally. Our result showing that decreasing exposure favours mixtures over sequences is also consistent with figure 3 from that paper [@Roush1998].

One consistent aspect of the results presented here can seem to contradict published advice and has been questioned when we have presented model results. Our model predicts that lower insecticide effectiveness or exposure will lead to slower spread of resistance in sole use and sequences (Figs 3 & 5, Table 2). This is consistent with expectations that lower effectiveness and exposure would lead to lower mortality of susceptibles and thus lower selection pressure for resistance [e.g. @Roush1998]. This can seem to contradict recommendations that poor implementation of insecticide interventions can promote the development of resistance. e.g. from [@FAO2012] "Certain pest control practices have consistently been shown to exacerbate the loss of susceptible pest populations and the development of resistance. These include: ... - the use of application rates that are below or above those recommended on the label, - poor coverage of the area being treated ...".

Poor implementation would likely cause lower effectiveness of and exposure to insecticides, so how can this increase the potential for resistance when our model suggests it would be decreased ? The explanation for this apparent difference is in the effect of dominance. Poor application can reduce the mortality of heterozygotes (RS) thus effectively increasing the dominance of the resistant allele and increasing its rate of evolution [@Barbosa2011][@FAO2012]. Our model results are consistent with this, showing that increasing dominance of restoration leads to faster spread of resistance (Fig 3C, 7). So, the effect of poor implementation of an insecticide intervention on resistance could be either positive or negative depending on the relative effects on dominance of the resistance gene, insecticide effectiveness and exposure. Similarly, the effects of poor application on the relative benefits of mixtures and sequences will depend on these trade-offs because of our result that effectivenes favours mixtures, exposure favours sequences and higher dominance favours neither (Table 4).

#### Implications
_todo get more of [@Roush1989] in here_

@@

_low exposure could be achieved by having active ingredients with some repellency._

_todo get Birget and Koella in & link to exposure_
'whereas repellency is beneficial to delay the evolution of resistance, other models have shown that it decreases the population-level protection of the insecticide.'

_todo, where to put this, earlier ?_
This model can explore when a mixture strategy may be favoured over a sequence purely for the evolution of resistance. Other factors also need to be considered. On the positive side a mixture is likely to kill more vectors and may also limit their ability to transmit the disease. On the negative side a mixture is more expensive.

_todo, somewhere earlier say good thing about effectiveness & exposure is that they can be modified_

#### Conclusions

We have developed flexible, modern model that can replicate previous modelling work and outputs that help us understand the mechanisms by which different insecticide application strategies may be favoured. For assessing the relative benefits of mixtures and sequences we have demonstrated the likely importance of insecticide effectiveness and exposure and costs of resistance. High insecticide effectiveness favours mixtures through the protective effect of one insecticide on the other. High exposure favours sequences through reducing the otherwise high selection pressures on both resistances together. High costs of resistance can favour mixtures because the selection pressures on both resistances are reduced. Similar processes will occur in other strategies such as rotations and mosaics and we are starting to investigate them. Predictions from the model can be tested against data from trials of new insectides to help advance guidelines on good practice.
